---
title: "10 Handy Tidyverse Tricks"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(palmerpenguins)
library(epiR)
library(broom)
library(medicaldata)
library(fuzzyjoin)
```

# 1. Putting it all in a frame

## Convert a vector to a dataframe

```{r}
num <- c("a", "b", "c", "d")
num
enframe(num)
```

## Named vectors are more useful

```{r}
names(num) <- c("x", "y", "z", "e")
enframe(num)
enframe(num, name = "letter", value = "number")
```

## A list column with dataframes

```{r}
dat1 <- mtcars
dat2 <- iris
dat3 <- penguins

l <- list(dat1, dat2, dat3)
names(l) <- c("mtcars", "iris", "penguins")

dat_l <- enframe(l, name = "data_name", "dataset")
dat_l
View(dat_l)
deframe(dat_l)
```

## A list column with lists

```{r}
dat <- penguins |> 
  select(bill_length_mm, bill_depth_mm, body_mass_g) |> 
  map(summary)

dat |> class()
length(dat)
enframe(dat) 
```

# 2. Counting

## Number of rows

```{r}
penguins |> 
  group_by(species) |> 
  summarise(n = n(), .groups = "drop")

penguins |> count(species)
```

## Take the sum

```{r}
penguins |> 
  group_by(island) |> 
  summarise(total_weight_g = sum(body_mass_g, na.rm = TRUE),
            .groups = "drop")

penguins |> count(island, wt = body_mass_g)
penguins |> count(island, wt = body_mass_g, name = "total_weight_g")
```

## Count the missing values

```{r}
penguins |> count(is.na(sex))
```


## Add a column of counts

```{r}
dat <- penguins |> 
  group_by(species) |> 
  mutate(n = n()) |> 
  ungroup()

View(dat)

dat1 <- penguins |> 
  add_count(species)

View(dat1)
```

# 3. Operate across() many columns

## mutate() to modify an existing column

Form 1

```{r}
dat <- penguins |> 
  mutate(
    across(bill_length_mm:body_mass_g,
           replace_na, 0),
    sex = fct_explicit_na(sex, na_level = "Unknown")
  )

View(dat)
```

Form 2

```{r}
dat <- penguins |> 
  mutate(
    across(bill_length_mm:body_mass_g,
           ~ replace_na(., 0)),
    sex = fct_explicit_na(sex, na_level = "Unknown")
  )

dat
```

## mutate() to create new columns

Form 1

```{r}
dat <- penguins |> 
  mutate(
    across(bill_length_mm:body_mass_g,
           list(new = replace_na), 0),
    sex = fct_explicit_na(sex, na_level = "Unknown")
  )

View(dat)
```

Form 2

```{r}
dat <- penguins |> 
  mutate(
    across(bill_length_mm:body_mass_g,
           list(new = ~ replace_na(., 0))),
    sex = fct_explicit_na(sex, na_level = "Unknown")
  )

View(dat)
```

## summarise() to aggregate columns

Form 1: use the same column names

```{r}
dat1 <- penguins |> 
  group_by(species, island) |> 
  summarise(
    across(bill_length_mm:body_mass_g,
           mean, na.rm = TRUE),
    .groups = "drop"
  )

dat1
```

Form 1: modify column names

```{r}
dat1 <- penguins |> 
  group_by(species, island) |> 
  summarise(
    across(bill_length_mm:body_mass_g,
           list(new = mean), na.rm = TRUE),
    .groups = "drop"
  )

dat1
```

Form 2: use the same column names

```{r}
dat1 <- penguins |> 
  group_by(species, island) |> 
  summarise(
    across(bill_length_mm:body_mass_g,
           ~ mean(., na.rm = TRUE)),
    .groups = "drop"
  )

dat1
```

Form 2: use the same column names

```{r}
dat1 <- penguins |> 
  group_by(species, island) |> 
  summarise(
    across(bill_length_mm:body_mass_g,
           list(new = ~ mean(., na.rm = TRUE))),
    .groups = "drop"
  )

dat1
```

# 4. List columns are awesome!

Form 1

```{r}
dat <- penguins |> 
  group_by(species) |> 
  nest() |> 
  rowwise() |> 
  mutate(
    mod = list(
      lm(body_mass_g ~ sex, data = data)
    ),
    pred = list(predict(mod))
  ) |> 
  ungroup()

dat

dat |> pull(mod)
```

Form 2

```{r}
dat <- penguins |> 
  nest_by(species) |> 
  mutate(
    mod = list(
      lm(body_mass_g ~ sex, data = data)
    ),
    pred = list(predict(mod))
  ) |> 
  ungroup()

dat

dat |> pull(mod)
```

# 5. Separate a column into pieces

## Prepare data

```{r}
dat <- tibble(
  name = paste(letters[1:6], letters[5:10], letters[11:16], letters[21:26], sep = "-"),
  value = 1:6
)

dat
```

## separate() into multiple columns

```{r}
dat |> 
  separate(col = name,
           into = c("first", "second", "third", "fourth"),
           sep = "-")

dat |> 
  separate(col = name,
           into = c("first", "second"),
           sep = "-",
           extra = "merge")

dat |> 
  separate(col = name,
           into = c("first", "second", "third"),
           sep = "-",
           extra = "merge")
```

## separate_rows() into multiple rows

```{r}
dat

new <- dat |> 
  separate_rows(name, sep = "-")

print(new, n = 25)
```

# 6. reduce() them to one

```{r}
v <- 1:4
v
sum(v)

1 + 2 + 3 + 4
reduce(v, `+`)
```

```{r}
gender <- tibble(id = letters[1:6], gender = sample(c("M", "F"), 6, TRUE))
age_cm <- tibble(id = letters[1:6], age = runif(6, min = 150, max = 190))
bmi <- tibble(id = letters[1:6], bmi = runif(6, min = 15, max = 40))

gender
age_cm
bmi

l <- list(gender, age_cm, bmi)
l

l |> reduce(left_join, by = "id")
```

# 7. Fuzzy join

## Prepare data

```{r}
dat1 <- tibble(
  district = c("Aberdeen City", "Aberdeenshire", 
               "Abertawe - Swansea", "Adur District", "Allerdale District (B)", 
               "Amber Valley District (B)", "Angus", "Argyll and Bute", "Arun District", 
               "Ashfield District", "Ashford District (B)", "Babergh District", 
               "Barking and Dagenham London Boro", "Barnet London Boro", "Barnsley District (B)",
                "Barrow-in-Furness District (B)", "Basildon District (B)",
               "Basingstoke and Deane District (B)", 
               "Bassetlaw District", "Bath and North East Somerset", "Bedford (B)", 
               "Bexley London Boro", "Birmingham District (B)", "Blaby District", 
               "Blackburn with Darwen (B)", "Blackpool (B)", "Blaenau Gwent - Blaenau Gwent", 
               "Bolsover District", "Bolton District (B)", "Boston District (B)", 
               "Bournemouth, Christchurch and Poole", "Bracknell Forest (B)", 
               "Bradford District (B)", "Braintree District", "Breckland District", 
               "Brent London Boro", "Brentwood District (B)",
               "Bro Morgannwg - the Vale of Glamorgan", 
               "Broadland District", "Bromley London Boro"),
  pop1 = c(8L,  2L, 14L, 14L, 14L, 14L, 3L, 5L, 14L, 14L, 14L, 14L, 14L, 14L, 
          14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 
          14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L)
)

dat2 <- tibble(
  district = c("Adur", "Allerdale", "Amber Valley", 
               "Arun", "Ashfield", "Ashford", "Babergh", "Barking and Dagenham", 
               "Barnet", "Barnsley", "Barrow-in-Furness", "Basildon", "Basingstoke and Deane", 
               "Bassetlaw", "Bath and North East Somerset", "Bedford", "Bexley", 
               "Birmingham", "Blaby", "Blackburn with Darwen", "Blackpool", 
               "Blaenau Gwent", "Bolsover", "Bolton", "Boston",
               "Bournemouth, Christchurch and Poole", 
               "Bracknell Forest", "Bradford", "Braintree", "Breckland", "Brent", 
               "Brentwood", "Bridgend", "Brighton and Hove", "Bristol, City of", 
               "Broadland", "Bromley", "Bromsgrove", "Broxbourne", "Broxtowe"
  ),
  pop2 = c(63869, 97527, 126678, 159827, 127151, 
          129281, 91401, 211998, 392140, 245199, 67137, 185862, 175729, 
          116839, 192106, 171623, 247258, 1141374, 100421, 148942, 139305, 
          69713, 79530, 285372, 69366, 395784, 121676, 537173, 151561, 
          139329, 330795, 76550, 144876, 290395, 463405, 129464, 331096, 
          98662, 96876, 113272)
)
```

## Join data

```{r}
dat1
dat2
library(fuzzyjoin)

dat_join <- stringdist_full_join(
  dat1, dat2,
  by = "district",
  method = "jw", distance_col = "dist"
) |> 
  arrange(district.x, dist) |> 
  group_by(district.x) |> 
  slice(1) |> 
  ungroup() |> 
  mutate(district.y = if_else(dist < 0.25, district.y, NA_character_)) 

dat_join
```

# 8. Permutations

```{r}
regimen <- LETTERS[1:5]
gender <- c("M", "F")
color <- c("red", "green", "blue")

dat1 <- crossing(regimen, gender)
print(dat1)

dat2 <- crossing(regimen, gender, color)
print(dat2, n = 30)

# Base version
expand.grid(regimen, gender)
expand.grid(regimen, gender, color)
```

# 9. One two punch

## Prepare data

```{r}
dat <- medicaldata::cytomegalovirus
glimpse(dat)

dat1 <- dat |> 
  mutate(
    across(c(cmv, prior.transplant), ~ if_else(. == 0, 2, .))
  ) |> 
  rename(tp = prior.transplant)
```

## Base R approach

```{r}
(data_table <- table(dat1$tp, dat1$cmv))
res <- epi.2by2(data_table)
res

res_sum <- summary(res)
res_sum$massoc.detail$RR.strata.wald
```

## Tidy approach

```{r}
data_table <- dat1 |> 
  select(tp, cmv) |> 
  table()

data_table

dat1 |> 
  select(tp, cmv) |> 
  table() |> 
  epi.2by2() |> 
  tidy() |> 
  filter(str_detect(term, "^RR"))
```

# 10. Visualising categoricla variables

## Prepare data

```{r}
dat <- ggplot2::mpg |> 
  rename(manu = manufacturer)

glimpse(dat)

dat |> count(manu)
```

## Plot all manufacturers

```{r}
dat_count <- dat |> 
  count(manu)

dat_count

ggplot(dat_count, aes(x = n, y = manu)) +
  geom_col() +
  theme_light()
```

## Lump small groups together

```{r}
dat1 <- dat |> 
  mutate(
    manu = factor(manu),
    manu = fct_lump_n(manu, 5)
  )

dat_count <- dat1 |> 
  count(manu)

ggplot(dat_count, aes(x = n, y = manu)) +
  geom_col() +
  theme_light()

ggplot(dat_count, aes(x = n, y = fct_reorder(manu, n))) +
  geom_col() +
  theme_light()

ggplot(dat_count, aes(x = n, y = fct_reorder(manu, n, .desc = TRUE))) +
  geom_col() +
  theme_light()
```
